---
title: "1. First steps in R and RStudio"
---

## Learning goals

After this lesson, you should be able to:

- explain the difference between R and RStudio;
- open an R project and run a script;
- create objects using `<-`;
- import a CSV file;
- inspect rows, columns, and variable types;
- find help when you do not remember a function.

::: {.research-question}
**Research question:** What information is contained in our employee survey, and
how is the dataset organized?
:::

## R and RStudio

**R** performs the calculations. **RStudio** is the interface in which we write,
run, organize, and inspect R code.

A useful beginner rule is:

> Write important work in a script, not only in the Console.

Console commands disappear easily. A saved script records what you did and can
be run again.

## Open the project

Open `r-for-management-research.Rproj`. Then open
`code/01-first-steps.R`.

## Load a package

```{r}
library(tidyverse)
```

A package adds functions to R. `tidyverse` contains tools for importing,
transforming, and visualizing data.

## Import the dataset

```{r}
employees <- read_csv(
  "data/employee_survey.csv",
  show_col_types = FALSE
)
```

Read this from right to left:

1. `read_csv()` reads the file;
2. the result is assigned to the object `employees`;
3. the object can be reused in later commands.

## Inspect the data

```{r}
head(employees)
glimpse(employees)
names(employees)
dim(employees)
```

- `head()` displays the first rows.
- `glimpse()` shows variables and their data types.
- `names()` lists the variable names.
- `dim()` gives the number of rows and columns.

## Understand common data types

| Type | Meaning | Example |
|---|---|---|
| `chr` | text or category | department |
| `dbl` | numeric value | performance |
| `int` | whole number | age |
| `lgl` | logical value | `TRUE` or `FALSE` |
| `date` | calendar date | campaign date |

A category may initially appear as text. We can convert it to a factor when
that helps the analysis.

## Select a variable

```{r}
employees$performance
```

The `$` operator selects one column from a data frame.

## Create your own objects

```{r}
course_name <- "Research Project Design"
number_of_employees <- nrow(employees)

course_name
number_of_employees
```

Object names should be descriptive and use underscores rather than spaces.

## Comments

```r
# This is a comment.
mean(employees$performance, na.rm = TRUE)
```

R ignores text after `#`. Comments explain why a step is being performed.

## Getting help

```{r}
#| eval: false
?mean
?read_csv
```

You can also place the cursor inside a function and press `F1` in RStudio.

::: {.common-mistake}
**Common mistake:** R is case-sensitive. `Performance`, `performance`, and
`PERFORMANCE` are different names.
:::

## Practice

1. Display the first ten rows with `head(employees, 10)`.
2. Find the number of columns using `ncol(employees)`.
3. Select the `department` variable.
4. Ask for help about `summary()`.

## Takeaway

An R analysis starts by creating a project, importing data into an object, and
checking its structure before calculating anything.
