---
title: "2. Clean, describe, and visualize data"
---

## Learning goals

You will learn to:

- inspect missing values;
- select, filter, recode, and summarize variables;
- calculate descriptive statistics;
- compare groups;
- create a histogram and a boxplot.

::: {.research-question}
**Research question:** What does the employee survey tell us about performance,
satisfaction, and differences between the training groups?
:::

## Import and clean names

```{r}
library(tidyverse)
library(janitor)

employees <- read_csv(
  "data/employee_survey.csv",
  show_col_types = FALSE
) |>
  clean_names()
```

The pipe `|>` sends the result of one step into the next step. It can be read as
“then”.

## Check missing values

```{r}
employees |>
  summarise(
    missing_performance = sum(is.na(performance)),
    missing_satisfaction = sum(is.na(job_satisfaction))
  )
```

`NA` means a value is missing. Missing data should be examined rather than
silently ignored.

## Select variables

```{r}
employees |>
  select(
    employee_id,
    department,
    training_group,
    performance
  )
```

## Filter rows

```{r}
employees |>
  filter(department == "Marketing")
```

Use `==` to ask whether two values are equal. A single `=` is normally used to
name a function argument.

## Create categories

```{r}
employees_clean <- employees |>
  mutate(
    remote_mode = case_when(
      remote_days == 0 ~ "On-site",
      remote_days <= 3 ~ "Hybrid",
      remote_days >= 4 ~ "Mostly remote"
    ),
    high_turnover_intention = if_else(
      turnover_intention >= 5,
      "High",
      "Not high"
    )
  )
```

`mutate()` creates or changes variables. `case_when()` is useful when a new
category depends on several rules.

## Count categories

```{r}
employees_clean |>
  count(department, sort = TRUE)
```

## Calculate descriptive statistics

```{r}
employees_clean |>
  summarise(
    n = n(),
    mean_performance = mean(performance, na.rm = TRUE),
    sd_performance = sd(performance, na.rm = TRUE),
    median_performance = median(performance, na.rm = TRUE),
    mean_satisfaction = mean(job_satisfaction, na.rm = TRUE)
  )
```

- The **mean** is the arithmetic average.
- The **median** is the middle value.
- The **standard deviation** describes spread around the mean.

`na.rm = TRUE` tells the function to remove missing values for that calculation.

## Compare groups

```{r}
employees_clean |>
  group_by(training_group) |>
  summarise(
    n = n(),
    mean_performance = mean(performance, na.rm = TRUE),
    sd_performance = sd(performance, na.rm = TRUE),
    .groups = "drop"
  )
```

`group_by()` changes the level at which a later summary is calculated.

## Visualize a distribution

```{r}
ggplot(employees_clean, aes(x = performance)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  labs(
    title = "Distribution of employee performance",
    x = "Performance score",
    y = "Number of employees"
  )
```

A histogram shows the shape, centre, and spread of a numeric variable.

## Compare distributions across groups

```{r}
ggplot(
  employees_clean,
  aes(x = training_group, y = performance)
) +
  geom_boxplot() +
  labs(
    title = "Performance by training group",
    x = NULL,
    y = "Performance score"
  )
```

A boxplot helps compare medians, spread, and unusual observations.

::: {.interpretation}
A visible difference in a graph is descriptive evidence. It does not by itself
show whether the difference is statistically distinguishable from random
variation, nor whether training caused the difference.
:::

## Practice

1. Calculate mean burnout by department.
2. Count employees in each `remote_mode`.
3. Draw a histogram of `job_satisfaction`.
4. Draw a boxplot of burnout by `remote_mode`.

## Common mistakes

::: {.common-mistake}
- Forgetting `na.rm = TRUE` when missing values are present.
- Treating a coded category as if it were a continuous number.
- Reporting only a mean without also checking the distribution.
- Using a graph to make a causal claim.
:::

## Takeaway

Cleaning and descriptive analysis come before formal tests. They reveal data
quality problems and help determine which method is appropriate.
