---
title: "3. Correlations, confidence intervals, and t-tests"
---

## Learning goals

You will learn to:

- interpret a correlation;
- run independent and paired t-tests;
- read a confidence interval and p-value;
- report an effect size;
- distinguish statistical evidence from practical importance.

::: {.research-question}
**Research questions:**
1. Are satisfaction and performance associated?
2. Do training and control groups differ?
3. Did participants improve after the intervention?
:::

## Correlation

```{r}
library(tidyverse)

employees <- read_csv(
  "data/employee_survey.csv",
  show_col_types = FALSE
)

cor.test(
  employees$job_satisfaction,
  employees$performance,
  use = "complete.obs"
)
```

Pearson's correlation, \(r\), ranges from -1 to +1:

- positive values indicate that larger values tend to occur together;
- negative values indicate that one variable tends to decrease as the other increases;
- values closer to zero indicate a weaker linear relationship.

A correlation does not establish causation.

## Independent-samples t-test

Use an independent t-test when comparing a numeric outcome between two separate
groups.

```{r}
independent_test <- t.test(
  performance ~ training_group,
  data = employees
)

independent_test
```

The formula means:

```text
numeric outcome ~ two-group variable
```

By default, R uses Welch's t-test, which does not require equal group
variances.

## Effect size

```{r}
library(effectsize)

cohens_d(
  performance ~ training_group,
  data = employees
)
```

A p-value addresses compatibility with a null model. An effect size addresses
the magnitude of the difference. Report both.

## Paired-samples t-test

Use a paired t-test when two measurements belong to the same participant.

```{r}
training <- read_csv(
  "data/training_experiment.csv",
  show_col_types = FALSE
)

paired_test <- t.test(
  training$post_score,
  training$pre_score,
  paired = TRUE
)

paired_test
```

The pairing matters because each post-score is linked to the same person's
pre-score.

## Difference in change between groups

```{r}
t.test(
  change_score ~ group,
  data = training
)
```

This asks whether the average improvement differs between training and control
participants.

## Confidence intervals

A 95% confidence interval gives a range of effect sizes compatible with the
data and model assumptions. A narrow interval is more precise than a wide
interval.

Do not interpret it as “there is a 95% probability that the true value is
inside this already calculated interval.”

## A reporting template

> The training group had a higher mean performance score than the control
> group. The estimated mean difference was **[difference]**, 95% CI
> **[lower, upper]**, Welch's \(t\)(**df**) = **value**, \(p\) = **value**,
> Cohen's \(d\) = **value**.

::: {.interpretation}
A small p-value does not automatically imply a large, useful, or causal effect.
Interpret the estimated difference, uncertainty, research design, and practical
context together.
:::

## Practice

1. Test whether burnout differs between managers and non-managers.
2. Calculate the correlation between workload and burnout.
3. Compare post-scores between the training and control groups.
4. Explain which test is paired and why.

## Common mistakes

::: {.common-mistake}
- Running a paired test on unrelated groups.
- Reporting only \(p < .05\).
- Saying “no effect” when a test is not significant.
- Claiming training caused a difference when group assignment was not random.
:::

## Takeaway

Choose the t-test based on the study design, then interpret the estimated
difference, confidence interval, effect size, and p-value together.
