---
title: "4. Reliability and scale construction"
---

## Learning goals

You will learn to:

- identify multi-item constructs;
- reverse-score negatively worded items;
- calculate Cronbach's alpha;
- inspect item diagnostics;
- construct a scale score with a transparent missing-data rule.

::: {.research-question}
**Research question:** Do the survey items consistently measure leadership and
engagement, and how can we combine them into usable variables?
:::

## Why use several items?

Management constructs such as leadership, engagement, trust, or perceived
support are not usually observed directly. Researchers often measure them with
several survey items.

A scale score is defensible only when the items are conceptually coherent and
show acceptable measurement properties.

## Reverse-coded items

Some items are worded in the opposite direction. In this dataset,
`leadership_4_reverse` and `engagement_4_reverse` are negative items.

For a 1–5 scale:

```text
reversed score = 6 - original score
```

```{r}
library(tidyverse)
library(psych)

employees <- read_csv(
  "data/employee_survey.csv",
  show_col_types = FALSE
) |>
  mutate(
    leadership_4 = 6 - leadership_4_reverse,
    engagement_4 = 6 - engagement_4_reverse
  )
```

If reverse coding is forgotten, the item can reduce reliability and distort
the scale.

## Select the items

```{r}
leadership_items <- employees |>
  select(
    leadership_1,
    leadership_2,
    leadership_3,
    leadership_4
  )

engagement_items <- employees |>
  select(
    engagement_1,
    engagement_2,
    engagement_3,
    engagement_4
  )
```

## Calculate Cronbach's alpha

```{r}
leadership_alpha <- alpha(leadership_items)
leadership_alpha
```

Focus on:

- raw alpha;
- corrected item-total correlations;
- alpha if an item is deleted;
- warnings about reverse direction.

Alpha is evidence about internal consistency, not proof that the scale is
unidimensional or valid.

## Build scale scores

We will require at least three answered items out of four.

```{r}
employees_scored <- employees |>
  mutate(
    leadership_valid_items = rowSums(!is.na(leadership_items)),
    engagement_valid_items = rowSums(!is.na(engagement_items)),
    leadership = if_else(
      leadership_valid_items >= 3,
      rowMeans(leadership_items, na.rm = TRUE),
      NA_real_
    ),
    engagement = if_else(
      engagement_valid_items >= 3,
      rowMeans(engagement_items, na.rm = TRUE),
      NA_real_
    )
  )
```

The missing-data rule should be decided and reported rather than improvised
after seeing the results.

## Describe the scales

```{r}
employees_scored |>
  summarise(
    mean_leadership = mean(leadership, na.rm = TRUE),
    sd_leadership = sd(leadership, na.rm = TRUE),
    mean_engagement = mean(engagement, na.rm = TRUE),
    sd_engagement = sd(engagement, na.rm = TRUE)
  )
```

## Interpretation checklist

Before using a scale, ask:

1. Do the items represent the same theoretical construct?
2. Were negative items coded in the same direction?
3. Are item-total relationships reasonable?
4. Does exploratory or confirmatory factor analysis support the structure?
5. Is the scale appropriate for this population and context?

::: {.common-mistake}
A high alpha can occur because items are repetitive. It does not establish
content validity, construct validity, or causality.
:::

## Practice

1. Calculate alpha for the four engagement items.
2. Calculate alpha before reverse-coding `engagement_4_reverse`.
3. Compare the result after correct reverse coding.
4. Create a `role_clarity` scale from its four items.

## Takeaway

Reliability analysis supports transparent scale construction, but it must be
combined with theory and evidence about dimensionality and validity.
