---
title: "5. Exploratory factor analysis and chi-square"
---

## Learning goals

You will learn to:

- explain what exploratory factor analysis is for;
- prepare survey items for EFA;
- inspect factorability;
- estimate and interpret a simple rotated factor solution;
- test association between two categorical variables with chi-square.

::: {.research-question}
**Research questions:**
1. Do the twelve survey items form the expected leadership, engagement, and
role-clarity dimensions?
2. Is remote-work mode associated with high turnover intention?
:::

## Exploratory factor analysis

EFA examines whether correlations among observed items can be represented by a
smaller number of latent dimensions.

It is exploratory: the pattern is learned from the data rather than fully
specified in advance.

## Prepare the items

```{r}
library(tidyverse)
library(psych)

employees <- read_csv(
  "data/employee_survey.csv",
  show_col_types = FALSE
) |>
  mutate(
    leadership_4 = 6 - leadership_4_reverse,
    engagement_4 = 6 - engagement_4_reverse
  )

survey_items <- employees |>
  select(
    leadership_1,
    leadership_2,
    leadership_3,
    leadership_4,
    engagement_1,
    engagement_2,
    engagement_3,
    engagement_4,
    role_clarity_1,
    role_clarity_2,
    role_clarity_3,
    role_clarity_4
  ) |>
  drop_na()
```

For this introductory exercise, rows with missing item responses are removed.
A real study should justify its missing-data strategy more carefully.

## Check factorability

```{r}
KMO(survey_items)
cortest.bartlett(
  cor(survey_items),
  n = nrow(survey_items)
)
```

- KMO assesses whether the correlation pattern is suitable for factor analysis.
- Bartlett's test examines whether the correlation matrix differs from an
identity matrix.

These diagnostics do not decide the factor solution on their own.

## Explore the number of factors

```{r}
fa.parallel(
  survey_items,
  fa = "fa"
)
```

Parallel analysis compares observed eigenvalues with those expected from random
data.

## Estimate three factors

```{r}
efa_model <- fa(
  survey_items,
  nfactors = 3,
  rotate = "oblimin",
  fm = "minres"
)

print(
  efa_model$loadings,
  cutoff = .30
)
```

We use an oblique rotation because management constructs can reasonably be
correlated.

A loading describes how strongly an item is related to a factor. Look for:

- several meaningful loadings on each factor;
- items loading strongly on their intended factor;
- limited cross-loadings;
- a solution that makes theoretical sense.

## Chi-square test

Create categorical variables:

```{r}
employees_categorical <- employees |>
  mutate(
    remote_mode = case_when(
      remote_days == 0 ~ "On-site",
      remote_days <= 3 ~ "Hybrid",
      remote_days >= 4 ~ "Mostly remote"
    ),
    high_turnover_intention = if_else(
      turnover_intention >= 5,
      "High",
      "Not high"
    )
  )
```

Build a contingency table:

```{r}
turnover_table <- table(
  employees_categorical$remote_mode,
  employees_categorical$high_turnover_intention
)

turnover_table
prop.table(turnover_table, margin = 1)
```

Run the test:

```{r}
chi_result <- chisq.test(turnover_table)

chi_result
chi_result$expected
```

The chi-square test asks whether the observed combination of categories differs
from what would be expected if the variables were independent.

::: {.interpretation}
A significant chi-square test indicates association, not the direction,
magnitude, mechanism, or causality. Row percentages and a suitable effect size
are needed for substantive interpretation.
:::

## Practice

1. Run EFA with two factors and compare the interpretability.
2. Identify the strongest-loading item for each factor.
3. Test association between department and managerial status.
4. Inspect expected cell counts before interpreting the chi-square result.

## Common mistakes

::: {.common-mistake}
- Choosing the number of factors only because it gives a preferred result.
- Treating PCA and common factor analysis as identical.
- Ignoring cross-loadings and theory.
- Interpreting chi-square as evidence that one category causes another.
:::

## Takeaway

EFA helps explore measurement structure; chi-square tests association between
categorical variables. Both require interpretation beyond a p-value.
