---
title: "6. Linear regression and diagnostics"
---

## Learning goals

You will learn to:

- translate a management question into a regression model;
- interpret coefficients, confidence intervals, and model fit;
- distinguish association from causation;
- inspect linearity, residuals, multicollinearity, and influential cases;
- communicate results without overstating them.

::: {.research-question}
**Research question:** Which marketing activities are associated with weekly sales, after accounting for the other activities and market conditions?
:::

## The idea

Linear regression estimates the expected change in a continuous outcome for a
one-unit change in a predictor, while holding the other included predictors
constant.

For this lesson:

- outcome: weekly sales in thousands of euros;
- predictors: search, social, and email advertising, discount, and competitor
  price index.

## Import and inspect

```{r}
library(tidyverse)
library(broom)
library(car)
library(performance)

marketing <- read_csv(
  "data/marketing_mix.csv",
  show_col_types = FALSE
)

glimpse(marketing)
summary(marketing)
```

## Begin with a graph

```{r}
marketing |>
  ggplot(aes(x = search_ads_k_eur, y = sales_k_eur)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Search advertising and weekly sales",
    x = "Search advertising (thousand EUR)",
    y = "Sales (thousand EUR)"
  )
```

A graph is not a substitute for a model, but it can reveal nonlinearity,
unusual observations, or data-entry problems.

## Estimate a multiple regression

```{r}
marketing_model <- lm(
  sales_k_eur ~
    search_ads_k_eur +
    social_ads_k_eur +
    email_ads_k_eur +
    discount_pct +
    competitor_price_index,
  data = marketing
)

summary(marketing_model)
```

A compact table is easier to reuse:

```{r}
tidy(marketing_model, conf.int = TRUE)
glance(marketing_model)
```

## Interpret one coefficient

Suppose the coefficient for `search_ads_k_eur` is positive. The interpretation
is:

> Holding the other variables in the model constant, an additional thousand
euros of search advertising is associated with an estimated change of **b**
thousand euros in weekly sales.

Do not say that advertising *caused* the change unless the research design
supports a causal conclusion.

## Model fit

`glance()` reports, among other statistics:

- **R-squared:** share of sample variation in sales explained by the model;
- **adjusted R-squared:** R-squared adjusted for the number of predictors;
- **F-test:** whether the model as a whole improves on an intercept-only model.

High R-squared does not prove that the model is correct or causal.

## Check multicollinearity

```{r}
vif(marketing_model)
```

Variance inflation factors indicate whether predictors contain strongly
overlapping information. They are a diagnostic signal, not an automatic rule
for deleting variables.

## Check assumptions visually

```{r}
check_model(marketing_model)
```

Focus on four questions:

1. Is the relationship approximately linear?
2. Are residuals distributed around zero without a clear pattern?
3. Is residual spread reasonably stable?
4. Are a few observations unusually influential?

## Inspect influential observations

```{r}
marketing_diagnostics <- augment(marketing_model) |>
  mutate(row_number = row_number()) |>
  arrange(desc(.cooksd))

marketing_diagnostics |>
  select(row_number, .fitted, .resid, .std.resid, .cooksd) |>
  slice_head(n = 5)
```

One week in the teaching data is intentionally unusual. Investigate it rather
than deleting it automatically. Ask whether it is an error, a genuine event,
or evidence that the model is incomplete.

::: {.interpretation}
A good regression report combines the research question, coefficient estimates,
uncertainty, model fit, diagnostics, and limitations. A p-value alone is not a
substantive conclusion.
:::

## Practice

1. Estimate a model with only the three advertising channels.
2. Add discount and competitor price index. What changes?
3. Identify the most influential week and inspect its raw values.
4. Write a three-sentence interpretation of one coefficient.

## Common mistakes

::: {.common-mistake}
- Interpreting every coefficient causally.
- Comparing coefficient sizes when variables use very different units.
- Ignoring diagnostic plots.
- Deleting unusual observations only because they reduce significance.
- Reporting R-squared without describing the outcome and sample.
:::

## Takeaway

Regression estimates conditional associations. Credible use requires both
substantive interpretation and diagnostic checking.
