---
title: "7. Moderation and mediation"
---

## Learning goals

You will learn to:

- distinguish moderation from mediation;
- create scale scores before modelling;
- estimate and visualize an interaction;
- estimate a simple statistical mediation model;
- explain why neither analysis automatically establishes causality.

## Two different questions

**Moderation asks when or for whom a relationship changes.**

> Is the association between workload and burnout weaker when organizational
support is high?

**Mediation asks through what statistical pathway variables are related.**

> Is leadership associated with performance partly through employee engagement?

These questions require different models.

## Prepare the constructs

```{r}
library(tidyverse)
library(broom)
library(interactions)
library(mediation)

employees <- read_csv(
  "data/employee_survey.csv",
  show_col_types = FALSE
) |>
  mutate(
    leadership_4 = 6 - leadership_4_reverse,
    engagement_4 = 6 - engagement_4_reverse,
    leadership = rowMeans(
      across(c(leadership_1, leadership_2, leadership_3, leadership_4)),
      na.rm = TRUE
    ),
    engagement = rowMeans(
      across(c(engagement_1, engagement_2, engagement_3, engagement_4)),
      na.rm = TRUE
    ),
    workload_c = workload - mean(workload, na.rm = TRUE),
    support_c = organizational_support -
      mean(organizational_support, na.rm = TRUE)
  )
```

Centering makes zero represent the sample mean. It can make the lower-order
coefficients easier to interpret, but it does not change the interaction test.

## Moderation: estimate an interaction

```{r}
moderation_model <- lm(
  burnout ~ workload_c * support_c + age + tenure_years,
  data = employees
)

tidy(moderation_model, conf.int = TRUE)
```

The `workload_c:support_c` term is the interaction. It tests whether the slope
of workload differs across levels of organizational support.

## Visualize the interaction

```{r}
interact_plot(
  moderation_model,
  pred = workload_c,
  modx = support_c,
  plot.points = TRUE,
  interval = TRUE
)
```

Interpret the pattern before focusing on significance:

- What happens to predicted burnout as workload rises?
- Is that increase different at lower and higher support?
- Are observations available across the plotted combinations?

::: {.interpretation}
When an interaction is included, the coefficient for workload is its estimated
slope when centered support equals zero—that is, at average support.
:::

## Mediation: estimate two regressions

First, create one complete-case dataset containing every variable used by
both models. This ensures that the mediator and outcome models use exactly the
same observations.

```{r}
mediation_data <- employees |>
  drop_na(leadership, engagement, performance, age, tenure_years)

nrow(mediation_data)
```

The mediator model predicts engagement:

```{r}
mediator_model <- lm(
  engagement ~ leadership + age + tenure_years,
  data = mediation_data
)
```

The outcome model predicts performance:

```{r}
outcome_model <- lm(
  performance ~ leadership + engagement + age + tenure_years,
  data = mediation_data
)
```

Estimate the indirect effect:

```{r}
set.seed(2026)
mediation_result <- mediate(
  mediator_model,
  outcome_model,
  treat = "leadership",
  mediator = "engagement",
  boot = TRUE,
  sims = 500
)

summary(mediation_result)
plot(mediation_result)
```

Key quantities include:

- average causal mediation effect reported by the software;
- average direct effect;
- total effect;
- proportion mediated.

The package uses causal terminology, but a cross-sectional observational dataset
does not, by itself, justify a causal mediation claim. In this lesson, describe
the result as a **statistical indirect association**.

## Practice

1. Remove the interaction and compare the model.
2. Plot the moderation without raw points. Which view is more informative?
3. Reverse the mediator and outcome conceptually. Why is theory essential?
4. Write separate one-paragraph interpretations for moderation and mediation.

## Common mistakes

::: {.common-mistake}
- Treating moderation and mediation as interchangeable.
- Interpreting lower-order terms without considering the interaction.
- Claiming a mechanism from cross-sectional correlations alone.
- Selecting a mediator only because the indirect effect is significant.
- Ignoring measurement quality of the constructs.
:::

## Takeaway

Moderation describes changing relationships; mediation describes an indirect
pathway. Both depend on theory, measurement, design, and cautious interpretation.
