---
title: "8. Logistic regression"
---

## Learning goals

You will learn to:

- recognize when a binary outcome requires logistic regression;
- estimate a beginner-friendly logistic model;
- interpret odds ratios cautiously;
- calculate predicted probabilities for meaningful profiles;
- distinguish classification from explanation.

::: {.research-question}
**Research question:** Which booking and guest characteristics are associated
with receiving a complimentary hotel-room upgrade?
:::

## Why not linear regression?

The outcome `upgrade` has two values:

- `0`: no upgrade;
- `1`: upgrade.

Logistic regression models the log-odds of the outcome and keeps predicted
probabilities between 0 and 1.

## Import and prepare

```{r}
library(tidyverse)
library(broom)

hotel <- read_csv(
  "data/hotel_upgrades.csv",
  show_col_types = FALSE
) |>
  mutate(
    loyalty_status = factor(
      loyalty_status,
      levels = c("None", "Silver", "Gold", "Platinum")
    ),
    special_event = factor(special_event),
    direct_booking = factor(direct_booking)
  )
```

Inspect the outcome:

```{r}
hotel |>
  count(upgrade) |>
  mutate(percent = 100 * n / sum(n))
```

## Estimate the model

```{r}
upgrade_model <- glm(
  upgrade ~
    loyalty_status +
    prior_stays +
    occupancy_rate +
    total_spend_eur +
    special_event +
    direct_booking,
  data = hotel,
  family = binomial
)

summary(upgrade_model)
```

## Convert coefficients to odds ratios

```{r}
tidy(
  upgrade_model,
  exponentiate = TRUE,
  conf.int = TRUE
)
```

An odds ratio:

- above 1 indicates higher estimated odds;
- below 1 indicates lower estimated odds;
- equal to 1 indicates no estimated change in odds.

Odds are not probabilities. An odds ratio of 2 does not mean the probability
doubles.

## Predicted probabilities

Predicted probabilities are often easier to communicate.

```{r}
guest_profiles <- tibble(
  loyalty_status = factor(
    c("None", "Gold"),
    levels = c("None", "Silver", "Gold", "Platinum")
  ),
  prior_stays = c(0, 7),
  occupancy_rate = c(90, 65),
  total_spend_eur = c(250, 900),
  special_event = factor(
    c("No", "No"),
    levels = levels(hotel$special_event)
  ),
  direct_booking = factor(
    c("No", "Yes"),
    levels = levels(hotel$direct_booking)
  )
)

guest_profiles |>
  mutate(
    predicted_probability = predict(
      upgrade_model,
      newdata = guest_profiles,
      type = "response"
    )
  )
```

The two profiles differ on several characteristics, so this comparison is a
scenario illustration rather than the isolated effect of one variable.

## A one-variable probability plot

```{r}
probability_data <- tibble(
  loyalty_status = factor(
    "Silver",
    levels = c("None", "Silver", "Gold", "Platinum")
  ),
  prior_stays = 0:12,
  occupancy_rate = mean(hotel$occupancy_rate),
  total_spend_eur = median(hotel$total_spend_eur),
  special_event = factor(
    "No", levels = levels(hotel$special_event)
  ),
  direct_booking = factor(
    "Yes", levels = levels(hotel$direct_booking)
  )
) |>
  mutate(
    probability = predict(
      upgrade_model,
      newdata = ., type = "response"
    )
  )

probability_data |>
  ggplot(aes(x = prior_stays, y = probability)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Predicted upgrade probability",
    x = "Prior stays",
    y = "Predicted probability"
  )
```

## Practice

1. Fit a smaller model using loyalty, occupancy, and direct booking.
2. Compare odds ratios with predicted probabilities.
3. Create two profiles that differ only in loyalty status.
4. Explain why a predictive association may not represent a fair decision rule.

## Common mistakes

::: {.common-mistake}
- Interpreting odds ratios as probabilities.
- Treating a significant predictor as a causal driver.
- Reporting accuracy without considering class balance.
- Using sensitive or proxy variables without ethical reflection.
- Assuming a simulated teaching model is suitable for deployment.
:::

## Takeaway

Logistic regression models binary outcomes. Predicted probabilities usually
provide the clearest applied interpretation, but ethical and design questions
remain essential.
