---
title: "Getting started"
---

## What you need

Install:

1. **R**, the statistical programming language;
2. **RStudio Desktop**, the interface used in the course;
3. **Quarto**, only if you want to render the complete website or reports.

You do not need previous programming experience.

## Download the course

The easiest beginner workflow is:

1. open the GitHub repository;
2. choose **Code → Download ZIP**;
3. extract the ZIP;
4. double-click `r-for-management-research.Rproj`.

Opening the `.Rproj` file sets the project folder as the working context. This
means that paths such as `data/employee_survey.csv` work consistently.

## Install the packages

Open `R/install_packages.R` and click **Source**, or run:

```{r}
#| eval: false
source("R/install_packages.R")
```

Then check the installation:

```{r}
#| eval: false
source("R/check_setup.R")
```

## Know the four RStudio panes

- **Source**: where you write and save scripts.
- **Console**: where R executes commands.
- **Environment**: objects currently stored in memory.
- **Files / Plots / Packages / Help**: files, visual output, packages, and documentation.

## Your first commands

```{r}
#| eval: false
2 + 2

course_name <- "Research Project Design"
number_of_students <- 30

course_name
number_of_students
```

`<-` assigns a value to an object. Read it as “gets”.

## Run code safely

Place the cursor on a line and use:

- Windows/Linux: `Ctrl + Enter`
- macOS: `Cmd + Enter`

Run code from the top of a script in order. If an object has not yet been
created, later code that depends on it will fail.

## File paths

Use project-relative paths:

```{r}
#| eval: false
employees <- readr::read_csv("data/employee_survey.csv")
```

Avoid paths tied to one computer, such as:

```r
# Do not use this:
"C:/Users/YourName/Desktop/course/data.csv"
```

## A simple debugging routine

When code fails:

1. read the first error message;
2. check spelling and capitalization;
3. check commas, brackets, and quotation marks;
4. confirm the package was loaded;
5. confirm the object and variable names exist;
6. run the previous lines again.

::: {.callout-tip}
Errors are normal. The useful question is not “Why am I bad at R?” but “What
specific information is the error giving me?”
:::
