---
title: "Quantitative Research Report"
author: "Student name"
date: today
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Research question

State one focused empirical question.

## Data and variables

Describe the sample, outcome, predictors, measurement scales, and missing-data
decisions. State whether the data are real or simulated.

```{r}
library(tidyverse)
library(broom)

data <- read_csv("data/employee_survey.csv", show_col_types = FALSE)
```

## Descriptive evidence

Present the observations needed to understand the sample and variables.

```{r}
glimpse(data)
summary(data)
```

## Analysis

Explain why the selected method matches the research question and variable
types. Run the analysis below.

```{r}
# Replace this example with your analysis.
model <- lm(performance ~ workload + organizational_support, data = data)
tidy(model, conf.int = TRUE)
glance(model)
```

## Interpretation

Explain direction, magnitude, uncertainty, and substantive meaning. Distinguish
association from causation.

## Diagnostics and robustness

Show the checks appropriate for the selected method and note any concerns.

## Limitations

Discuss design, measurement, sample, missing data, and generalizability.

## Reproducibility note

List the data file, script or report file, packages, and any random seed used.
